Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-03 paper.Rnw
Updated text to match WA xval figures, and reworked discussion

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
The previous experiment assesses \mb{} peformance on whole hippocampus
segmentation. In this experiment, we assess \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images using a leave-one-out
cross-validation (LOOCV) design.

@@

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is a validation approach in which an
algorithm is given all but one item in a dataset as training data (in our case,
atlas images and labels) and then is applied to the left-out item. This is done,
in turn, for each item in the dataset and the output across all items is
evaluated.

In this experiment, the high-resolution $0.3mm^3$ voxel Winterburn atlases are
used as the \mb{} atlas library, but in each round of LOOCV a $0.9mm^3$ voxel
target image corresponding to the left-out atlas is used so as to evaluate \mb{}
on standard 3T T1-weighted resolution images.

Specifically, this experiments has two parts, each using different sets of
$0.9mm^3$ voxel target images: trilinear subsampled versions of the Winterburn
atlas images (Experiment 2a), and separately acquired T1 BRAVO acquisitions of
four of the five Winterburn atlas subjects (Experiment 2b).  Image subsampling
of the Winterburn atlas images was performed using trilinear subsampling
techniques.

The 3T T1 BRAVO acquisitions of four of the five Winterburn atlas subjects were
separately obtained within a short time of the original atlas image
acquisitions.  Images were acquired on a 3T GE Discovery MR 750 system (General
Electric, Milwaukee, WI) using an 8-channel head coil with the enhanced fast
gradient recalled echo 3-dimensional acquisition protocol, EFGRE-BRAVO, with the
following parameters: $TE/TR/TI = 3.0ms/6.7ms/650ms$, flip angle $=8^\circ$,
$FOV=15.8cm$, slice thickness$=0.9mm$, 176 in-plane steps for an approximate
isotropic resolution of 0.9mm dimension voxels.

The template library is composed of all $0.9mm^3$ voxel target images (either
the {\em BRAVO} or {\em subsampled} sets), plus an additional set of 16 of
healthy subjects 3T T1 images acquired separately (Table
\ref{tab:WAval-healthy-demographics}). These images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution of
0.9mm dimension voxels.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used. Figure \ref{fig:LOOCV-schematic}
illustrates schematically the experimental set up.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figure/WinterburnSubfieldLOOCVSchematic}
  \caption{A schematic illustration of the experimental set up of the
  leave-one-out cross-validation (LOOCV) subfield experiment.}
  \label{fig:LOOCV-schematic}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
<<WAval-prep>>=
wa.xval.vols = na.omit(read.csv('data/WAval.csv'))
wa.xval.casted = dcast(melt(wa.xval.vols, id.vars=c("file","version")), file + variable ~ version)
wa.xval.casted = rename(wa.xval.casted, c("variable"="region"))
levels(wa.xval.casted$region) <-list("CA1"="X1", "CA1"="X101", 
                            "Subiculum"="X2","Subiculum"="X102", 
                            "CA4/DG"="X4","CA4/DG"="X104", 
                            "CA2/CA3"="X5","CA2/CA3"="X105", 
                            "SR/SL/SM"="X6","SR/SL/SM"="X106")

#d = ddply(na.omit(wa.xval.casted), c('region'), function (df) {
#  icc = ICC(df[4:5])
#  data.frame(
#    icc  = icc$results$ICC[3],
#    f    = icc$results$F[3],
#    lower= icc$results$"lower bound"[3],
#    upper= icc$results$"upper bound"[3],
#  )
#})

fmt_tstat = function(a, b) {
  test = t.test(a,b, paired=T)
  p = sprintf("%0.3f", max(0.001,round(test$p.value,3)))
  t = sprintf("%0.3f", round(test$statistic,3))
  paste(t," (",p,")")
}

wa.p = ddply(wa.xval.casted, c("region"), function (df) {
  with(df, {
       data.frame(
         res = fmt_tstat(gold_0.3mm, gold_0.9mm),
         bra = fmt_tstat(gold_0.3mm, mb_bravo),
         mbr = fmt_tstat(gold_0.3mm, mb_0.9mm)
       )})
  })

wa.p = rename(wa.p, c(
  'region'='Subregion','res'='Resampled',
  'bra'='MAGeT on BRAVO','mbr'='MAGeT on Resampled'))

wa.xval.error = ddply(wa.xval.casted, c("file", "region"), function(df) {
  with(df, 
  data.frame(
    resampled = (gold_0.9mm - gold_0.3mm) / gold_0.3mm * 100, 
    mb_bravo  = (mb_bravo   - gold_0.3mm) / gold_0.3mm * 100, 
    mb_0.9mm  = (mb_0.9mm   - gold_0.3mm) / gold_0.3mm * 100))})

wa.s = ddply(wa.xval.error, c('region'), function(df) {
  with(df,data.frame(
    m_res = mean(resampled,na.rm=T), sd_res = sd(resampled,na.rm=T),
    m_bravo = mean(mb_bravo,na.rm=T), sd_bravo = sd(mb_bravo,na.rm=T),
    m_mb_res = mean(mb_0.9mm,na.rm=T), sd_mb_res = sd(mb_0.9mm,na.rm=T)))})
@

\begin{figure}[h!]
<<fig:WAxval-kappa>>=
wa.n.sim = read.csv('data/winterburn.xval.similarity.csv')
wa.n.sim$label = as.factor(wa.n.sim$label)
levels(wa.n.sim$label) <-list(
  "CA1"=1, "CA1"=101, "Subiculum"=2,"Subiculum"=102, 
  "CA4/DG"=4,"CA4/DG"=104, "CA2/CA3"=5,"CA2/CA3"=105, 
  "SR/SL/SM"=6,"SR/SL/SM"=106)

# kappa vs. region
ggplot(wa.n.sim, aes(x=as.factor(label),y=k)) + 
  geom_boxplot(fill='lightblue') + 
  xlab("Subregion") + 
  ylab("Dice's Similarity Coefficient (DSC)")
@
  \caption{Similarity of \mb{} segmentation of subregions and resampled
  Winterburn atlas labels at $0.9mm^3$ voxel resolution over all folds of a
  leave-one-out cross validation.}
\end{figure}

\begin{figure}[h!]
<<fig:WAxval-Bland-Altman>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv')
winterburn = read.csv('data/winterburn.volumes.csv')

wa.n.vol.melt = melt(wa.n.vol, id.vars=c('atlases','subject','res'))
winterburn.melt = melt(winterburn, id.vars=c('subject','res'))
winterburn.cast = dcast(winterburn.melt, subject + variable ~ res)
wa.n.all.vols = merge(wa.n.vol.melt, winterburn.cast, by=c('subject','variable'))
#wa.n.vols.mean = ddply(wa.n.all.vols,c('subject','variable'),function (df) {
#  data.frame(
#    mb=mean(df$value),
#    X.3=mean(df$'0.3'),
#    X.9=mean(df$'0.9'))})

@@

wa.n.xval.bland.limits = ddply(wa.n.all.vols,c("variable"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})
ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and \\mb subregion volume ' (mm^3))) + 
  ylab(expression('Difference between manual and \\mb subregion volume ' (mm^3))) + 
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{Bland-Altman plots across all folds of Winterburn atlas cross validation.}
\end{figure}

In this experiment the \mb{} segmentation of hippocampal subfields is assessed
through a leave-one-out validation in which resampled images ($0.9mm^3$
voxels) of each Winterburn atlas subject are segmented using the remaining
high-resolution Winterburn atlas subjects' images.

Figure \ref{fig:WAval-vol-boxplot} shows the mean relative percent error in volume, 
with respect to the full-resolution Winterburn atlas, of each hippocampal subfield
of the resampled Winterburn labels, and of the \mb{} segmentations of the resampled
Winterburn atlas images and T1 BRAVO Winterburn subject acquisitions.  A
paired-samples t-test was conducted to compare the distribution of relative
percent error in the \mb{} segmentation volumes and the resampled segmentation
volumes. There was no significant difference found for the CA1, Subiculum, and
CA2/CA3 subregions. The CA4/DG subregion shows significantly lower percent error 
  ($M_{resampled}=$\Sexpr{wa.s$m_res[wa.s$region=='CA4/DG']},
  $SD_{resampled}=$\Sexpr{wa.s$sd_res[wa.s$region=='CA4/DG']};
  $M_{MAGeT(BRAVO)}=$\Sexpr{wa.s$m_bravo[wa.s$region=='CA4/DG']},    
  $SD_{MAGeT(BRAVO)}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='CA4/DG']};    
  $M_{MAGeT(resampled)}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='CA4/DG']},    
  $SD_{MAGeT(resampled)}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='CA4/DG']};    
  $p < 0.001$), and the SR/SL/SM shows significantly different percent error in the opposite
direction 
  ($M_{resampled}=$\Sexpr{wa.s$m_res[wa.s$region=='SR/SL/SM']},
  $SD_{resampled}=$\Sexpr{wa.s$sd_res[wa.s$region=='SR/SL/SM']};
  $M_{MAGeT(BRAVO)}=$\Sexpr{wa.s$m_bravo[wa.s$region=='SR/SL/SM']},    
  $SD_{MAGeT(BRAVO)}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='SR/SL/SM']};    
  $M_{MAGeT(resampled)}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='SR/SL/SM']},    
  $SD_{MAGeT(resampled)}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='SR/SL/SM']};    
  $p < 0.001$). Bland-Altman plots comparing \mb{} to subsampled segmentation
volume reveals a proportional bias towards underestimating volumes in all
subregions, and prominant fixed biases in the volumes of the CA4/DG and SR/SL/SM subregions 
congruent with significant differences seen in percent error in volume
associated with subsampling alone.

Figure \ref{fig:subfield-montage} shows slices subfield segmentations for a
single subject for qualitative inspection.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WAxval - t.test comparing Manual volumes and Computed volumes of subregions 
%          across folds
<<tab:WAxval-t-test-manual-computed, echo=F, results="asis",cache=FALSE>>=
latex(wa.p, file="", size="scriptsize",landscape=FALSE, ctable=TRUE,
    caption="\\textbf{Paired-sample t-tests comparing manually segmented
             Winterburn atlas subregion volumes and computed volumes (reported
             as $t (p)$). }", 
    col.just=c("l","r","r","r"),
    caption.loc = c("bottom"), rowname=NULL, 
    label="tab:WAxval-t-test-manual-computed")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<fig:WAval-vol-boxplot>>=
wa.xval.error.melted = melt(wa.xval.error,id.vars=c("file", "region"))
wa.xval.error.melted = rename(wa.xval.error.melted, c("variable"="measure", "value"="percenterr"))
levels(wa.xval.error.melted$measure) <- list("Subsampled atlas labels"="resampled",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

ggplot(wa.xval.error.melted, aes(y=percenterr,x=measure,colour=measure)) +
  facet_grid(.~region) + 
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() + 
  geom_point(group=1, position="dodge") + 
  labs(y="Percent error in volume", x="Subfield") +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  scale_colour_discrete("") + 
  theme(legend.direction = "vertical", legend.position = "bottom")
@
  \caption{{\bf Percent error in computed unilateral hippocampal subfield
segmentation volume relative to manual segmentation volume.} 
  Percent error is measured against the volumes of the unmodified Winterburn
  atlas subfield segmentations.
  {\bf Subsampled atlas labels} volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$ voxels.  
  {\bf MAGeT on subsampled atlas images} volumes are \mb{} segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT on T1 BRAVO} volumes are \mb{} segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.\label{fig:WAval-vol-boxplot}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: bland-altman 
\begin{figure}[h!]
<<fig:WAval-bland>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm', 'gold_0.9mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.volumes$diff = wa.xval.volumes$gold_0.9mm - wa.xval.volumes$volume 
wa.xval.volumes$mean = (wa.xval.volumes$gold_0.9mm + wa.xval.volumes$volume)/2
wa.xval.bland.limits = ddply(wa.xval.volumes,c("region"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})

ggplot(wa.xval.volumes, aes(x=mean, y=diff, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and \\mb subregion volume ' (mm^3))) + 
  ylab(expression('Difference between manual and \\mb subregion volume ' (mm^3))) + 
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))

@
  \caption{{\bf Bland-Altman plots comparing unilateral \mb{} subfield
segmentation volume with resampled subfield segmentation volume.} The overall
mean difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by
dashed horizontal lines. Linear fit lines are shown for each target image type.
Note, points below the mean difference indicate overestimation of the volume
with respect to the resampled volume, and vice versa. 
  \label{fig:WAval-bland}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                Discussion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tabulation of segmentation accuracies of existing methods (ADNI)
%
<<ADNI1-literature, echo=F, results="asis",cache=FALSE>>=
# caption = 
#   "\\textbf{Automated segmentation accuracy (overlap with manual labels) of the 
#    ADNI dataset.} For each method, the number of manually labelled atlases used for 
#    training, the best Dice's overlap measure, the disease classes measured, and 
#    the validation procedure are shown. LOOCV = Leave-one-out cross-validation. 
#    Some studies of automated segmentation of ADNI images are excluded because they 
#    do not not provide overlap measures for the hippocampus \\cite{Heckemann2011, Chupin2009}.
#   "
# 
# tab = read.csv("data/ADNI-existing-work.csv")
# tab$Notes <- NULL
# 
# adni_lit = subset(tab, Dataset=="ADNI")
# adni_lit$Dataset <- NULL
# 
# 
# latex(adni_lit, #file="", 
#       size="scriptsize",
#       caption=caption, 
#       rowname=NULL,
#       col.just=c("p{2in}","p{1in}","p{1in}","p{1in}","p{1in}"),
#       title="")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The core claim the \mb{} method is based on, that we can meaningfully bootstrap
a template library from a small set of labelled atlas images, is validated in
the cross validation conducted in Experiment 1 (and the replication in
Experiment 5, Supplementary Materials). We find that both increasing the number
of atlases and the number of templates used improves \mb{} segmentation over and
above basic-\multiatlas segmentations using the same number of atlas images.
That is, by taking the extra step of generating a template library from a small
atlas set \mb{} is able to improve the final segmentation similarity. The
magnitude of this improvement is greatest with a small number of atlases, but
even with larger atlas libraries we have found that generating a template
library reduces the variability in segmentation precision (i.e.  \mb{} more
consistently produces high quality segmentations than does basic-\multiatlas{}
segmentation). These effects do not appear dependant on the hippocampal
segmentation protocol used.

@@

Finally, we have provide evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,VanLeemput2009}. While recent work demonstrates that
subfield segmentations can be used for classification of AD, MCI, and CN, there
has been no explicit validation of the segmentations being produced other than
indirectly via classification accuracy (for instance, the semi-automated
subfield segmentations produced in \citet{Yushkevich2009}).  Our work
demonstrates that we can reliably reproduce segmentations, corresponding in volumes 
to the input atlases used, with only modest error for the CA1, subiculum, and 
CA4/dentate subregions.

The fact that CA2/CA3 and molecular layers cannot be reasonably reproduced
should not be surprising as these are extremely thin and spatially convoluted
regions that originally required high-resolution MRI for identification and so
it is likely that the extents of these regions are well below the resolution and
contrast offered by standard T1-weighted images. In addition, some manual
subfield segmentation methodologies do not attempt to parse these regions even
on high-resolution scans. See Table \ref{ab:subfield-segmentation-protocols} for
a comparison of manual subfield segmentation methodologies. Further complicating
this, is that that different research groups have different operational
definitions for these different subfields. The disagreement in the community has
led to an international working group devoted to normalizing the ontology and
segmentation rules for the hippocampal subfields
(\url{http://www.hippocampalsubfields.com/}). The definitional and operational
disagreements make evaluating the accuracy of automated methodologies with
respect to a "ground truth" (e.g. using overlap similarity metrics, such as
Dice's Kappa) not particularly meaningful until these disagreements have been
resolved.  Thus, because of this, and the inability to discern subregions in
standard T1-weighted images, we opted to speak of the precision or reliability
of \mb{} subfield segmentation rather than speak of accuracy.

In conclusion, we have presented a flexible \multiatlas{}-based framework that has
considerable advantages over other methods since only a small number of atlases
is required to initialize the algorithm.  We demonstrate that our method works
robustly over hippocampal definitions, different disease populations, and
different acquisition types. Finally, we also demonstrate that accurate
identification of the hippocampal subfields is possible.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
The previous experiment assesses \mb{} peformance on whole hippocampus
segmentation. In this experiment, we aim to evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9m^3$-isotropic voxels.  We
use a modified leave-one-out cross-validation (LOOCV) design.

@@

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is a validation approach in which an
algorithm is given all but one item in a dataset as training data (in our case,
atlas images and labels) and then the algorithm is applied to the left-out item.
This is done, in turn, for each item in the dataset and the output across all
items is evaluated together.

In this experiment, the Winterburn atlases are resampled to $0.9mm^3$-isotropic
voxel images to simulate standard 3T T1-weighted resolution images. Image
subsampling is performed using trilinear subsampling techniques.

In each round of LOOCV, one atlas is selected and segmented with \mb{} once
using each possible triplet of atlas images from the remaining four unselected
atlases. We chose to use an atlas library with an odd number of images so as to
ensure an unbiased label fusion when using majority voting (see Discussion).

The template library used has a total of 19 images composed of all $0.9mm^3$
voxel target images plus an additional set of 14 of healthy subject 3T T1 images
acquired separately (Table \ref{tab:WAval-healthy-demographics}). These images
were acquired on a 3T GE Discovery MR 750 system (General Electric, Milwaukee,
WI) using an 8-channel head coil with the enhanced fast gradient recalled echo
3-dimensional acquisition protocol, EFGRE-BRAVO, with the following parameters:
$TE/TR/TI = 3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution of
0.9mm dimension voxels.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}

\begin{figure}
<<fig:WAxval-kappa>>=
wa.n.sim = read.csv('data/winterburn.xval.similarity.csv')
wa.n.sim$label = as.factor(wa.n.sim$label)
levels(wa.n.sim$label) <-list(
  "CA1"=1, "CA1"=101, "Subiculum"=2,"Subiculum"=102, 
  "CA4/DG"=4,"CA4/DG"=104, "CA2/CA3"=5,"CA2/CA3"=105, 
  "SR/SL/SM"=6,"SR/SL/SM"=106)

# kappa vs. region
ggplot(wa.n.sim, aes(x=as.factor(label),y=k)) + 
  geom_boxplot(fill='lightblue') + 
  xlab("Subregion") + 
  ylab("Dice's Similarity Coefficient (DSC)")

CA1   = subset(wa.n.sim,label=='CA1')$k
CA23  = subset(wa.n.sim,label=='CA2/CA3')$k
CA4DG = subset(wa.n.sim,label=='CA4/DG')$k
Sub   = subset(wa.n.sim,label=='Subiculum')$k
SRLM  = subset(wa.n.sim,label=='SR/SL/SM')$k
@
  \caption{Similarity of \mb{} segmentation of subregions and the resampled
  Winterburn atlas segmentations at $0.9mm^3$ voxel resolution, over all folds of a
  leave-one-out cross validation. Overlap score for each hemisphere is measured
  separately. \label{fig:WAxval-kappa}}
\end{figure}

In this experiment the \mb{} segmentation of hippocampal subfields is assessed
through a leave-one-out validation in which resampled images ($0.9mm^3$
voxels) of each Winterburn atlas subject are segmented using the remaining
high-resolution Winterburn atlas subjects' images as atlases.

Figure \ref{fig:WAval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subregion across all subjects and folds of the validation. Mean (standard
deviation) DSC scores of the subregions are: 
  \Sexpr{mean(CA1)}   ($\pm$\Sexpr{sd(CA1)}) for the CA1 region,  
  \Sexpr{mean(CA23)}  ($\pm$\Sexpr{sd(CA23)}) for the CA2/CA3 regions,  
  \Sexpr{mean(CA4DG)} ($\pm$\Sexpr{sd(CA4DG)}) for the CA4/DG regions,  
  \Sexpr{mean(Sub)}   ($\pm$\Sexpr{sd(Sub)}) for the Subiculum region, and
  \Sexpr{mean(SRLM)}  ($\pm$\Sexpr{sd(SRLM)}) for the moleculare regions.
Bland-Altman plots comparing \mb{} to resampled subregion volumes reveal a
proportional bias towards underestimating volumes in all subregions (Figure
\ref{fig:WAxval-Bland-Altman}). Prominent fixed biases in the volumes of the
CA4/DG and SR/SL/SM subregions are congruent with the lower similarity scores
seen in these regions.

\begin{figure}
<<fig:WAxval-Bland-Altman>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv')
winterburn = read.csv('data/winterburn.volumes.csv')

wa.n.vol.melt = melt(wa.n.vol, id.vars=c('atlases','subject','res'))
winterburn.melt = melt(winterburn, id.vars=c('subject','res'))
winterburn.cast = dcast(winterburn.melt, subject + variable ~ res)
wa.n.all.vols = merge(wa.n.vol.melt, winterburn.cast, by=c('subject','variable'))

@@

wa.n.xval.bland.limits = ddply(wa.n.all.vols,c("variable"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})
ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and \\mb subregion volume ' (mm^3))) + 
  ylab(expression('Difference between manual and \\mb subregion volume ' (mm^3))) + 
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{\textbf{Bland-Altman plots comparing unilateral hippocampal subregion volume
across all folds of Winterburn atlas cross validation.} The overall mean
difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by dashed
horizontal lines. Linear fit lines are shown. Note, points below the mean
difference indication overestimation of the volume with respect to the resampled
volume, and vice versa. \label{fig:WAxval-Bland-Altman}}
\end{figure}

Figure \ref{fig:subfield-montage} shows slices subfield segmentations for a
single subject for qualitative inspection.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                Discussion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in Experiment 1 (and the replication in
Experiment 5, Supplementary Materials). We find that both increasing the number
of atlases and the number of templates used improves \mb{} segmentation over and
above basic-\multiatlas segmentations using the same number of atlas images.
That is, by taking the extra step of generating a template library from a small
atlas set \mb{} is able to improve the final segmentation similarity. The
magnitude of this improvement is greatest with a small number of atlases, but
even with larger atlas libraries we have found that generating a template
library reduces the variability in segmentation precision (i.e.  \mb{} more
consistently produces high quality segmentations than does basic-\multiatlas{}
segmentation). These effects do not appear dependant on the hippocampal
segmentation protocol used.

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,VanLeemput2009}. While recent work demonstrates that
subfield segmentations can be used for classification of AD, MCI, and CN, there
has been no explicit validation of the segmentations being produced other than
indirectly via classification accuracy (for instance, the semi-automated
subfield segmentations produced in \citet{Yushkevich2009}).  Our work
demonstrates that we can reliably reproduce segmentations for the CA1,
subiculum, and CA4/dentate subregions.  That the CA2/CA3 and molecular
layers are less well reproduced should not be surprising as these are extremely
thin and spatially convoluted regions that originally required high-resolution
MRI for identification and so it is likely that the extents of these regions are
well below the resolution and contrast offered by standard T1-weighted images.

This points to a larger issue of how to truly validate subfield segmentations,
both in high resolution images and in standard T1-weighted images.  There are
several manual subfield segmentation methodologies, and they do not agree on
which regions can be differentiated, even on high-resolution scans. See Table
\ref{ab:subfield-segmentation-protocols} for a comparison of manual subfield
segmentation methodologies. A further complication is that different
researchers have differing operational definitions for the subfields and how
they ought to be parcellated. The disagreement in the community has led to an
international working group devoted to normalizing the ontology and segmentation
rules for the hippocampal subfields
(\url{http://www.hippocampalsubfields.com/}). The definitional and operational
disagreements make evaluating the accuracy of automated methodologies with
respect to a "ground truth" (e.g. using overlap similarity metrics, such as
Dice's Kappa) not particularly meaningful until these disagreements have been
resolved.  Thus, because of these disagreements and the difficulting discerning
subregions in standard T1-weighted images, we opted to speak of the precision or
reliability of \mb{} subfield segmentation rather than speak of accuracy.

In conclusion, we have presented a flexible \multiatlas{}-based framework that has
considerable advantages over other methods since only a small number of atlases
is required to initialize the algorithm. We demonstrate that our method works
robustly over hippocampal definitions, different disease populations, and
different acquisition types. Finally, we also demonstrate that accurate
identification of the hippocampal subfields is possible. 
% TODO: say that we have shown the viability of template generation, and as such
% it may also serve as a useful addition to existing segmentation techniques
